---
title: "Assignment 4"
author: "Brian Yoo"
date: '2022-11-01'
output: html_document
---

```{r echo = FALSE, eval = TRUE, message = FALSE, include = FALSE}
source("../source/analysis.R")
```

### Abstract 

This report recalls the dataset that contains information on the United State jails. The dataset contains the different kinds of population in jail and other counts/rates related to the US prison.

### Introduction 

The data from this report can reveal patterns of inequality. The total imprisonment value of certain groups of people may proportionately differ and there may be a significant gap between certain groups who are susceptible to imprisonment. It also tackles the issue of incarceration. In certain years, there are more imprisonment in certain locations. It’s important to understand why this happened and if there is a solution that can be established from the dataset. The dataset provides more than just information on jail counts. It provides other essential details such as the deaths in custody reporting program (DCRP), which helps identify if there is any disproportionality within the population. The direct stakeholders may be the people who are imprisoned; the report from this data is directly focused on them and the attention may impact their treatment in the future. The indirect stakeholders include reporters using this dataset to update news on this topic, law enforcement who need to be cautious of who they arrest, and the US government court who makes the final decision on someone’s imprisonment.

### Data Summary 

#### Which state has the highest population of black people in jail?

`r state_highest_black` has the highest black prison population in the U.S.

#### Which year has the highest number of population in jail?

The year `r highest_pop_year` has the highest number of population in jail.

#### Which county has the highest DCRP?

`r dcrp_county` is the county that has the highest DCRP.

These three values helps us learn about the patterns of inequality within the 
United States. The variable "state_highest_black" shows us which state has the highest population of black people in jail within the United States. The 
variable "highest_pop_year" shows us which year has the highest number of
people in jail. Lastly, the variable "dcrp_county" shows us which county has
the highest rate of death in custody reporting program within the United States.

### Growth of the U.S. Prison Population

``` {r, echo=FALSE}
get_year_jail_pop <- function() {
  result <- trend %>% 
    group_by(year) %>% 
    filter(total_jail_pop != "NA") %>% 
    summarise(total_jail_pop = sum(total_jail_pop)) 
  return(result)
}

plot_year_jail_pop_us<- function() {
  ggplot(get_year_jail_pop(), aes(x = year, y = total_jail_pop)) + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Increase of Jail Population in U.S. (1970-2018)", 
       x = "Years", 
       y = "Total Jail Population")
}

plot_year_jail_pop_us()
```

A key pattern shown from this graph is that the number of total people
imprisoned in jail has increased exponentially in recent years. Starting from 
1980 to 2010, the number of people in prison has gone up by around 600,000, and
it's starting to lessen after 2010.

### Growth of Prison Population by State 

``` {r, echo=FALSE}

get_jail_pop_by_states <- function(states) {
  result <- trend %>% 
    filter(str_detect(state, paste(states, collapse = "|")) == TRUE) %>% 
    group_by(year, state) %>% 
    filter(total_jail_pop != "NA") %>% 
    summarise(total_jail_pop = sum(total_jail_pop)) 
  return(result)
} 

plot_jail_pop_by_states <- function(states) {
  ggplot(get_jail_pop_by_states(states), aes(x = year, y = total_jail_pop)) + 
    geom_line(aes(color = state, linetype = state)) +
    labs(title = "Jail Population in States (1970-2018)", caption = "Jail population for three different states", 
         x = "Years", 
         y = "Total Jail Population")
}
  
plot_jail_pop_by_states(c("WA", "OR", "CA"))
```

This chart shows us a visualization of how the rate of jail population in certain states differ. In this chart there is a comparison between the state of
California, Oregon, and Washington. The chart shows California had an exponential increase from 1980 to 1990, while Oregon and Washington had a 
steady linear increase from 1970 to 2018.


### Variable Comparison that Reveals Potential Patterns of Inequality

``` {r, echo=FALSE}
comparison <- function() {
  result <- trend %>%
    select(year, white_jail_pop, black_jail_pop) %>%
    gather(key = "variable", value = "value", -year, na.rm = TRUE)
  return(result)
}

plot_comparison <- function() {
  ggplot(comparison(), aes(x = year, y = value)) + 
    geom_line(aes(color = variable, linetype = variable)) + 
    scale_color_manual(values = c("darkred", "steelblue")) +
    labs(title = "White vs Black Jail Population in U.S. (1970-2018)", 
         x = "Years", 
         y = "Total Jail Population")
}

plot_comparison()
```

This chart shows the comparison between the total population of black
people in jail vs the total population of white people in jail. The chart
shows that the black population has significantly more number, than the white
population. The white population is represented by dotted blue lines, while
the black population is represented by red solid lines.

### A Map that Shows Potential Patterns of Inequality that Vary Geographically

``` {r, echo=FALSE}
library(maps)

black_jail_data <- trend %>% 
  select(state, year, black_jail_pop) %>% 
  filter(year == 2018) %>% 
  group_by(state) %>% 
  summarise(black_jail_pop = sum(black_jail_pop, na.rm = TRUE)) %>% 
  mutate(state_name = tolower(state.name[match(state, state.abb)]))

map_state <- map_data("state") %>% 
 rename(state_name = region) %>% 
  left_join(black_jail_data, by= "state_name")

jail_state <- function () {
  result <- map_state %>% 
    filter(!is.na(map_state$black_jail_pop))
  return(result)
}

map1 <- function() {
  ggplot(jail_state(), aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = black_jail_pop), color = "black")
}

plot_map2 <- function() {
  map1() + scale_fill_gradient(name = "Population Count",
                               low = "yellow", high = "red", na.value = "grey50") +
    labs(title = "Black Jail Population in U.S. (2018)") +
    theme(
      axis.text.x= element_blank(),        
      axis.text.y = element_blank(),       
      axis.ticks = element_blank(),       
      axis.title.y = element_blank(),
      axis.title.x = element_blank(),
      rect = element_blank())
} 
plot_map2()
```

This map shows the 2018 jail count for the Black population in the 
United States.In the map, we can see that the states with higher jail
counts for Black groups are colored in red and the states with lower 
count is colored in yellow. It's noticeable that the southern states 
have higher counts.